---
title: "Cyclistic bike-share data analysis"
author: "Ilya Semenov"
date: "2022-10-06"
output: html_document
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(xml2)
library(tidyverse)
library(stringr)
library(readr)
library(lubridate)
library(ggplot2)
library(scales)
```

# Problem statement

Based on a [Coursera problem statement document](https://d3c33hcgiwev3.cloudfront.net/aacF81H_TsWnBfNR_x7FIg_36299b28fa0c4a5aba836111daad12f1_DAC8-Case-Study-1.pdf?Expires=1665187200&Signature=H0BPQYIvO2gT32qdFKVqFz4UiOFIwhEsfGOVIptLicPlmiVpAOIiQ3BdI-~pmtiI1cTC9cNjQVVWBU7C8QNDiNDftDL4GvEIR12-3mtK9IhMi3R3fyWiT-Ocjl~ZDNhsoFclPAUicrnI6i0uxh7B3wREe4qXY9vsmowZHwgLctM_&Key-Pair-Id=APKAJLTNE6QMUY6HBC5A) find differences in a bike usage between annual members and casual riders.
Key stakeholder: Lily Moreno, Cyclistic marketing director.

# Summary

Compared with casual users annual members prefer bike-share usage for **work commuting**. It means preference towards shorter trips with duration 15 minutes and less and more active service usage during weekdays.

# Data preparation

### Data import
Cyclistic bike-usage data obtained by the following [link](https://divvy-tripdata.s3.amazonaws.com/index.html). Data are credible to Cyclistic exclusively. No signs of bias present. Data are anonymous.
Data import sequence for years 2021-2022 is as follows:

```{r include=TRUE, echo=TRUE, message=FALSE, warning=FALSE}
fpath <- 'https://divvy-tripdata.s3.amazonaws.com/'
get_filename <- function(x) {
  x$Key[[1]][[1]]
}

files_list <- as_list(read_xml(fpath))
filenames_list <- unlist(sapply(files_list$ListBucketResult, get_filename))

trips <- data.frame()
for (filename in filenames_list) {
  if (str_detect(filename, 'zip') & (str_detect(filename, '2021') | str_detect(filename, '2022'))) { # Get data for 2021 and 2022 only
    temp <- tempfile()
    download.file(paste0(fpath, filename), temp)
    trips <- rbind(trips, read_csv(temp))
    unlink(temp)
  }
}
```

### Data filtering and preprocessing

Data filtering included following transformations:

* only last 365 days of the bike-share service usage data were taken into account
* in order to correctly calculate trip duration only records with valid (not NA) started_at and ended_at fields were left
* trips with duration less than 30 seconds or more than 10 hours were filtered
* column with trip duration was added
* column with trip duration bin was added

The code is as follows:

```{r include=TRUE, echo=TRUE, message=FALSE, warning=FALSE}
trips_clean <- trips %>% 
  mutate(trip_duration = as.numeric(difftime(ended_at, started_at))) %>% 
  filter(!is.na(trip_duration)) %>% 
  filter(trip_duration > 29 & trip_duration < 60*60*10) %>%    # drop trips less than 30 seconds and more than 10 hours
  filter(started_at > max(started_at) - days(365)) %>%  # last 365 days only due to memory limits
  mutate(trip_duration_bin = case_when(
    trip_duration < 300 ~ '1: < 5 mins',
    trip_duration >= 300 & trip_duration < 900 ~ '2: 5-15 mins',
    trip_duration >= 900 & trip_duration < 1800 ~ '3: 15-30 mins',
    trip_duration >= 1800 & trip_duration < 3600 ~ '4: 30-60 mins',
    trip_duration >= 3600 & trip_duration < 7200 ~ '5: 1-2 hours',
    trip_duration >= 7200 ~ '6: > 2 hours'
  )) # set bins for trip durations

rm(trips) # clean memory
```

### Data preparation and processing step resume

The only datasource used for the analysis is the Cyclistic bike-usage dataset.

# Data analysis

### Weekly usage

Casual users tend to use the service more on holidays while annual members prefer service usage on weekdays. It's suggests the hypothesis of annual members use the service fo **work commuting**.
Total weekly usage is as follows:

```{r include=TRUE, echo=TRUE, message=TRUE, warning=FALSE}
table(wday(trips_clean$started_at, label = TRUE, locale = 'en_US.UTF8'), trips_clean$member_casual)
```

### Trip duration preferences, members vs. casual

```{r include=TRUE, echo=TRUE, message=TRUE, warning=FALSE}
ggplot(data = trips_clean, aes(x = wday(started_at, label = TRUE, locale = 'en_US.UTF8'), fill = trip_duration_bin)) +
  geom_bar() +
  facet_wrap(~member_casual) +
  scale_y_continuous(labels = number) +
  xlab('') +
  ylab('Rides count') +
  labs(caption = 'Source: Cyclistic bike-share dataset, last 365 days usage')
```

Annual members prefer short trips on weekdays, one more argument in favor of **commuting** hypothesis.

### Trip duration preferences by vehicle type

```{r include=TRUE, echo=TRUE, message=TRUE, warning=FALSE}
ggplot(data = trips_clean, aes(x = trip_duration_bin)) + 
  geom_histogram(stat = 'count') +
  facet_wrap(~member_casual + rideable_type) +
  theme(axis.text.x = element_text(angle = 90)) +
  scale_y_continuous(labels = number) +
  xlab('Trip duration bin') +
  ylab('Rides') +
  labs(title = 'Number of rides by duration and rider type', caption = 'Source: Cyclistic bike-share dataset, last 365 days usage')
```

Annual members preference towards short rides is valid both for classic and electric bikes. **Commuting** support argument #3.

# Recommendations

According to the analysis the marketing recommendations for Cyclicist are as follows:

* focus on casual users who often make short trips (30 minutes and less) on weekdays, propose annual subscription
* design a special annual program for casual electric bike usage since the service consumption pattern for electric bikes is very similar for both casual users and annual members
* consider special offer design for casual users who like to take long trips (more than 1 hour) on holydays
